Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2019-08-16 WholePlotTrajectories.tex
manuscript final

\Abstract{
Anticipate the impact of the global change context on tropical forests
is now urgent. This calls to highlight the taxonomic and functional
facets of tree community post-disturbance trajectories and the
underlying processes. It also calls to clarify the link between
biodiversity and disturbance, specifically the scope of the Intermediate
Disturbance Hypothesis (IDH) that is debated in tropical forests. In
this study, we analyzed the tree community trajectories following a
disturbance gradient in a Neotropical forest over 30 years. We
considered community taxonomic and functional diversity, composition and
redundancy trajectories. We based on the annual botanical inventories of
75 ha of a Neotropical forest and on large trait datasets comprising 7
leaf, stem and life-history traits. We highlighted a decoupling between
taxonomic trajectories, that differed among communities, and functional
trajectories, remained similar and convergent. We explained this
decoupling by the variations in community functional redundancy
mitigating the functional impact of disturbance. The humped-shaped
taxonomic trajectories which amplitude varied with disturbance intensity
validated the IDH (Intermediate Disturbance Hypothesis). The IDH was
however dismissed for functional diversity. Although consistent, the
recovery of community composition, diversity and redundancy remained
unachieved after 30 years. This acknowledged the need of decades-long
cycles without disturbance to ensure a complete recovery, and questioned
tropical forest community resilience after repeated disturbances.
}

@@

\begin{document}

@@

The large areas covered with tropical forests worldwide hold crucial
environmental, economic and social values. They provide wood and
multiple non-timber forest products, shelter a diversified fauna, and
ensure cultural and human well-being. They regulate as well the local
and regional climates, and the carbon, water and nutrient cycles.
However, the growing demand in forests products together with current
global changes increase the pressure on remaining natural undisturbed
forests \citep{Morales-Hidalgo2015}. These threats affect the
disturbance regime that naturally shapes and maintains the structure,
composition and functioning of tree communities
\citep{Schnitzer2001, Anderson-Teixeira2013, Sist2015}. To anticipate
the fate of tropical forests in this context, understanding the response
of communities to disturbance and deciphering the underlying ecological
processes is of utmost importance. Disturbance impacts forest
communities in changing both abiotic and biotic environments, through
modifications in the fluxes of light, heat and water
\citep{Goulamoussene2017}. Although the forest cover is generally
maintained after disturbance, there is a succession of changes in the
structure and functioning of ecological communities that defines a
post-disturbance ecosystem trajectory. For now, these trajectories have
been largely studied through trajectories of forest structural
parameters such as aboveground biomass, tree height or stem density
\citep{Piponiot2016, Rutishauser2016}. Recent works have identified some
of the determinants of post-disturbance biomass trajectories, like
pre-disturbance forest structure and composition or post-disturbance
environmental parameters \citep{Herault2018}. Regarding tree community
diversity and composition however, post-disturbance trajectories have
not been as thoroughly understood \citep{Guitet2018, Molino2001}. Given
the variety of species response to disturbance and the huge diversity of
tropical forests, manifold post-disturbance biodiversity trajectories
might emerge \citep{Lindenmayer2012, Garcia_florez2017}.

An early conceptual basis of the linkage between biodiversity and
disturbance is the Intermediate Disturbance Hypothesis (IDH). The IDH
assumes a relationship between community diversity and the intensity and
frequency of disturbance events, stating that community diversity peaks
at intermediate level of disturbance \citep{Connell1978}. The theory
assumes that the environmental fluctuations following disturbance foster
both competitively superior species and fast colonizers and thus prevent
competitive exclusion \citep{Shea2004, Pulsford2016}. In tropical
forests, however, observations of the IDH often diverge from theoretical
expectations \citep{Hugues2007, Sheil2003, Norden2017}, and the
validation of the theory remains controversial
\citep{Hubbell2001, Fox2013, Sheil2013}. In tropical forests the
processes underlying the IDH might indeed be complicated by the huge
diversity found in tropical tree communities and the variety of
ecological processes that may underlie the IDH (colonization,
facilitation, coexistence, etc)
\citep{Lindenmayer2012, Garcia_florez2017}. It is then crucial to test
the IDH in this context and determine its scope in time, using ecosystem
time-explicit trajectories \citep{Sheil2003, Shea2004}.

Analysing community response to disturbance requiers a set of metrics,
to grasp all aspects of community changes
\citep{Sheil2003, Shea2004, Mayfield2010}. The analysis should first
consider community composition, which is crucial for conservation issues
and which reveals the pool of species fostered (or hampered) by
disturbance \citep{Lavorel2002, Bellwood2006}. Then, diversity metrics
encompassing both richness and evenness should also be considered to
assess the changes in community abundance distribution. Besides,
functional approaches have been shown to usefully complement pure
taxonomic approaches because they shed light on the species biological
attributes that directly link community diversity, composition and
redundancy to ecosystem functioning \citep{Violle2007b, Baraloto2012a}.
In that respect, a vast literature allowed recognizing major traits that
represent species ecological strategy and determine how they respond to
changing conditions \citep{Diaz2005}. Specifically in tropical forests,
the functional approach revealed the post-disturbance deterministic
processes entailing a shift from a dominance of ``conservative''
slow-growing species dealing with scarce resources, to a dominance of
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \citep{Rees2001, Reich2014, Herault2011}. This shift
is translated into the trajectories of average community value of key
functional traits related to resource acquisition (leaf and stem traits)
and life-history strategy (seed mass, maximum size)
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}.

The functional approach also encompasses the analysis of functional
redundancy, that quantifies the amount of shared trait values among
species \citep{Carmona2016}. The typical high functional redundancy of
hyper-diverse tropical forests \citep{Bellwood2006} mitigates the
impacts of species removal on ecosystem functioning and determines
communities resilience after disturbance \citep{Elmqvist2003, Diaz2005}.

In this study, we monitored over 30 years the response of 75 ha of
Neotropical forest plots set up on a gradient of disturbance intensity,
from 10 to 60\% of above-ground biomass (AGB) loss. We made use of a
large functional traits database encompassing major leaf, stem and
life-history traits in order to draw the taxonomic and functional
trajectories in terms of richness, evenness, composition and redundancy.
Specifically, (i) we draw taxonomic and functional post-disturbance
trajectories and examine the underlying ecological process, (ii) we
discuss the scope of the IDH regarding the different facets of community
diversity and the time after disturbance, and (iii) we analyze community
resilience and time to recovery.

@@

Paracou station in French Guiana (5\textdegree 18'N and
52\textdegree 53'W) is located in a lowland tropical rain forest in a
tropical wet climate with mean annual temperature of 26\textdegree C,
mean annual precipitation averaging 2980 mm.y\textsuperscript{-1} (30-y
period) and a 3-month dry season (\textless{} 100
mm.month\textsuperscript{-1}) from mid-August to mid-November, and a
one-month dry season in March \citep{Wagner2011}. Elevation ranges from
5 to 50 m. Across all plots the topography mainly corresponds to
hilltops or hillsides, while bottomlands cover less than 1 \% of the
area. Plots are shallow ferralitic acrisols over a layer of transformed
saprolite with low permeability and lateral drainage. Soil conditions
are homogeneous, to the exception of the highest hilltops where the
thick surface allows a free vertical drainage
\citep{Gourlet-Fleury2004}.

The experiment is a network of twelve 6.25 ha plots (Table
\ref{tab:Tab1}) that underwent three disturbance treatments in 1987
according to a randomized plot design \citep{Gourlet-Fleury2004}.\\
The experiment comprised three replicates of three sylvicultural
treatments (hereafter plots T1, T2 and T3) and three control plots (T0).
All treatments T1, T2 and T3 comprised the logging of 10 trees/ha with
50 cm minimum DBH that belonged to a set of 58 commercial species
\citep{Gourlet-Fleury2004}. Treatment T2 additionally comprised a
thinning treatment by poison-girdling of non-commercial, randomly
selected species with an average 30 trees/ha with 40 cm minimum DBH.
Treatment T3 additionally comprised the logging of 15 trees /ha with 40
cm minimum DBH and the poison-gidling of 20 trees/ha with a 50 cm
minimum DBH, all belonging to non-commercial species. Considering the
silvicultural treatments and the following damage, disturbance intensity
was measured as the percentage of aboveground biomass (\%AGB) lost
between the first inventory in 1984 and five years after disturbance
\citep{Piponiot2016} estimated with the BIOMASS R package
\citep{Rejou2017}. The three treatments were then transformed into a
continuous disturbance intensity gradient with increasing of
above-ground biomass (AGB) lost.

@@

\caption{\label{tab:Tab1}Intervention table, summary of the disturbance intensity for the 4 plot treatments in Paracou. Treatment intensities are defined by the minimum logging DBH (Diameter at Breast Height), the type of logged species (commercial or not), the density of logged trees, and the total AGB (Above Ground Biomass) lost after treatment.}
\centering
\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
\toprule
Treatment & Timber & Thinning & Fuelwood & \%AGB lost\\
\midrule
Control & - & - & - & 0\\
T1, low & DBH $\geq$ 50 cm, commercial species, $\approx$ 10   $trees.ha^{-1}$ & - & - & $[12-33]$\\
T2, intermediate & DBH $\geq$ 50 cm, commercial species, $\approx$ 10  $trees.ha^{-1}$ & DBH $\geq$ 40 cm, non-valuable species, $\approx$ 30   $trees.ha^{-1}$ & - & $[33-56]$\\
T3, high & DBH $\geq$ 50 cm, commercial species, $\approx$ 10  $trees.ha^{-1}$ & DBH $\geq$ 50 cm, non-valuable species, $\approx$ 15  $trees.ha^{-1}$ & 40 cm $\leq$ DBH $\leq$ 50 cm, non-valuable species,\ $\approx$ 15 $trees.ha^{-1}$ & $[35-56]$\\
\bottomrule
\end{tabu}
\end{table}

@@

The study site corresponds to a tropical rainforest typical of the
Guiana Shield with a dominance of \emph{Fabaceae},
\emph{Chrysobalanaceae}, \emph{Lecythidaceae} and \emph{Sapotaceae}. In
the 12 experimental plots, all trees above 10 cm DBH have been mapped
and measured annually since 1984. Trees are first identified with a
vernacular name assigned by the forest worker team, and afterward with a
scientific name assigned by botanists during regular botanical
campaigns. In 1984, specific vernacular names were given to 62
commercial or common species whereas more infrequent ones were
identified under general identifiers only distinguishing trees and
palms. From 2003, botanical campaigns have been conducted every 5 to 6
years to identify all trees at the species level but identification
levels still varied among plots and campaigns.

This variability of protocols in time raised methodological issues as
vernacular names usually correspond to different botanical species. This
resulted in significant taxonomic uncertainty that had to be propagated
to composition and diversity metrics. The uncertainty propagation was
done through a Bayesian framework reconstituting complete inventories at
genus level from real incomplete ones on the basis of
vernacular/botanical names association. Vernacular names were replaced
through multinomial trials based on the association probability
\(\big[\alpha_1, \alpha_2,..., \alpha_V\big]\) observed across all
inventories between each vernacular name \emph{v} and all species
\(\big[s_1, s_2,..., s_N\big]\):

@@

Six functional traits representing leaf economics (leaf thickness,
toughness, total chlorophyll content and specific leaf area) and stem
economics (wood specific gravity and bark thickness) came from the
BRIDGE project \footnote{http://www.ecofog.gf/Bridge/}. Trait values
were assessed from a selection of individuals located in nine permanent
plots in French Guiana, including two in Paracou, and comprised 294
species pertaining to 157 genera. Whenever a species was in the dataset
but missed some trait values (10\% of the species), missing values were
filled using multivariate imputation by chained equation
\citep{Mice2011}. To account for the phylogenetic signal in the filling
process, imputations based on samples from the same genus or from the
same family. Whenever a species was not in the dataset, it was
attributed a set of trait values randomly sampled among species of the
next higher taxonomic level (same genus or family). Two life-history
traits (maximum specific height and seed mass) came from the Mariwenn
database \footnote{https://www.ecofog.gf/mariwenn/}. The database
compiles information from a vast literature on the flora of French
Guiana \citep{Ollivier2007} and comprises 362 species pertaining to 188
genera. As seed mass information was classified into classes, no data
filling process was applied and analyses were restricted to the
botanical species recorded.

@@

Because of the variability in the precision of botanical identification
efforts, we were constrained to conduct the taxonomic composition and
diversity analysis at the genus level. Taxonomic and functional
trajectories of community composition were followed in a two-dimensional
NMDS ordination plane. Two NMDS using abundance-based (Bray-Curtis)
dissimilarity measures were conducted to map either taxonomic or
functional composition, the latter based on the seven leaf, stem and
life history traits (without seed mass classes). Trajectories along time
were reported through the Euclidean distance between the target
inventories and the reference pre-disturbance inventories of 1984.
Univariate trajectories of the leaf, stem and life-history traits were
also visualized with the community weighted means (CWM)
\citep{Diaz2007}. Species seed mass were given in 5 mass classes, and
seed mass trajectories were reported as the proportion of each class in
the inventories (Supplementary materials).

The taxonomic and functional trajectories were analysed from the
reference pre-disturbance inventories of 1984. The taxonomic diversity
was reported through species richness and the Hill number translation of
the Simpson index \citep{Hill1973}. These metrics allowed assessing the
taxonomic richness as well as evenness, through the comparison between
these two metrics: results will be discussed directly in terms of
taxonomic richness and evenness. Both indices belong to the set of HCDT
or generalized entropy, respectively corresponding to the 0 and 2 order
of diversity (q), recommended for diversity studies \citep{Marcon2015}.
The functional diversity was reported using the functional richness and
functional evenness, through the Rao index of quadratic entropy. The Rao
index combines species abundance distribution, and the average pairwise
functional dissimilarity between species computed by the Gower distance.

The impacts of the initial disturbance levels on the maximum gain or
loss in richness and evenness were tested with the Spearman rank
correlation tests. Richness and evenness trajectories were besides
analyzed through polynomial regression between (i) taxonomic and
functional richness and evenness and (ii) the initial \%AGB loss at 10,
20 and 30 years after disturbance.

Finally, functional redundancy was measured as the overlap among species
in community functional space \citep{Carmona2016}. First, the
individuals of the trait database were mapped in the plane of the first
two axes from a PCA analysis, which lowered the weight of correlations
among traits as the PCA axes were combinations of most decoupled traits.
For each species, the traits probability density (TPD) were computed
from the mapping of individuals through two-dimension kernel density
estimators. Second, for each community, the TPD weigthed by species
abundance were summed accross the functional space. Third, the
functional space was divided into a 100 x 100 grid and the number of
species with a positive TPD was counted in each cell. The average count
across cells minus 1 returned the Community Functional Redundancy, which
was the average number of species in the community that share the same
trait values.

@@

From 1984, the first pre-disturbance inventory, to 2015 (28 years after
disturbance), 828 388 individual trees and 591 botanical species
pertaining to 223 genera and 64 families were recorded.

While both taxonomic and functional composition remained stable in
undisturbed communities (Fig. \ref{fig:NMDSplans}), they followed marked
and consistent trajectories in disturbed plots over time.

The functional composition trajectory in disturbed communities resembled
to cyclic compositional changes with an incomplete recovery to the
initial composition (Fig.\ref{fig:NMDSplans}). The maximum dissimilarity
with the initial state was positively correlated with the disturbance
intensity for both taxonomic and functional composition
(\(\rho_{Spearman}^{Taxonomic}=0.87\) and
\(\rho_{Spearman}^{Functional}=0.90\) respectively). The maximum
dissimilarity with the initial state was reached for taxonomic
composition between 15 to 25 years and around 20 to 25 years for
functional composition.

@@

\caption{Plot trajectories in terms of taxonomic composition (\textbf{(a)} and \textbf{(c)}) and functional composition (\textbf{(b)} and \textbf{(d)}) in a two-dimensional NMDS plane. Lower panels (\textbf{(c)} and \textbf{(d)}) represent the Euclidean distance to initial condition along the 30 sampled years. Shaded areas are the credibility intervals.}\label{fig:NMDSplans}
\end{figure*}

In control plots, Community Weighted Means (CWM) of functional traits
remain stable in time.\\
In disturbed plots, they mostly followed unimodal trajectories, either
stabilizing or returning towards their initial values, to the exception
of leaf chlorophyll content, which continued to increase 30 years after
disturbance for some highly disturbed plots. Maximum height at adult
stage (\emph{Hmax}), leaf toughness and wood specific gravity
(\emph{WSG}) decreased in time and then slightly increased, but remained
significantly lower than their initial value (Fig. \ref{fig:CWM}). Bark
thickness and specific leaf area (\emph{SLA}) both increased in time.
Bark thickness remained substantially high after 30 years, and
\emph{SLA} had almost recovered to its initial value. Whatever the
functional traits, the maximum difference to initial value was highly
correlated to the disturbance intensity. Positive correlations were
observed for Leaf thickness, chlorophyll content, SLA and bark thickness
(\(\rho_{Spearman}^{Leaf thickness}=0.76\),
\(\rho_{Spearman}^{Chlorophyll content}=0.60\),
\(\rho_{Spearman}^{SLA}=0.93\),
\(\rho_{Spearman}^{Bark thickness}=0.71\)). Negative correlation was
observed for Leaf toughness, WSG and Hmax
(\(\rho_{Spearman}^{Leaf toughness}=-0.53\),
\(\rho_{Spearman}^{WSG}=-0.75\), \(\rho_{Spearman}^{Hmax}=-0.40\)) The
proportions of the three lightest seed mass classes increased in all
disturbed plots. After 30 years the proportion of lightest seed mass
class decreased while it stabilized for the two other lightest seed mass
classes (Supp. Mat. - Fig. S2).

@@

AM, EM \& BH designed the study, developed the analysis framework and
interpreted the results. AM wrote the manuscript with contributions by
EM \& BH. All authors gave final approval for publication.

>>>>

\Abstract{
In the current global change context, it is urgent to anticipate the
fate of tropical forests. This means understanding tree community
response to disturbance and the underlying processes. In that respect,
we aim here to clarify taxonomic and functional post-disturbance
trajectories, and determine the scope of the Intermediate Disturbance
Hypothesis (IDH) that remains debated in tropical forests. We analyzed
community trajectories following a disturbance gradient in a Neotropical
forest over 30 years. We considered trajectories along time of community
taxonomic and functional trajectories in terms of richness, evenness,
composition, and redundancy. We based on the annual botanical
inventories of 75 ha of a Neotropical forest and on large trait datasets
comprising seven leaf, stem, and life-history traits. We identified a
decoupling between taxonomic composition, differing among communities,
and functional composition, remaining similar and convergent. The
taxonomic diversity followed humped-shaped trajectories depending on the
disturbance intensity, which validated the IDH (Intermediate Disturbance
Hypothesis). The functional diversity trajectories,however, were
homogeneous among all plots and dismissed the IDH. We explained this
decoupling by the variations in community functional redundancy that
mitigated the functional impact of disturbance. Although consistent, the
recovery of community composition, diversity, and redundancy remained
unachieved after 30 years. These results acknowledged the need of
decades-long cycles without disturbance to ensure community complete
recovery, and questioned community resilience after repeated
disturbances.
}

@@

\usepackage{amsthm}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}{Lemma}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\newtheorem{corollary}{Corollary}[section]
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{definition}
\newtheorem{exercise}{Exercise}[section]
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\begin{document}

@@

The large areas covered with tropical forests worldwide hold crucial
environmental, economic, and social values. They provide wood and
multiple non-timber forest products, shelter a diversified fauna, and
ensure cultural and human well-being. They regulate local and regional
climates, as well as carbon, water and nutrient cycles. However, the
growing demand in forests products together with current global changes
increase the pressure on remaining undisturbed forests
\citep{Morales-Hidalgo2015}. These threats affect the natural
disturbance regime that defines and maintains the structure,
composition, and functioning of tree communities
\citep{Schnitzer2001, Anderson-Teixeira2013, Sist2015}. To anticipate
the fate of tropical forests, it is urgent to understand tree community
response to disturbance, and the underlying ecological processes. The
forest cover is generally maintained following disturbance, but
modifications in the fluxes of light, heat, and water
\citep{Goulamoussene2017} change community abiotic and biotic
environments. These changes translate into post-disturbance community
trajectories that have been largely studied through trajectories of
forest structural parameters such as aboveground biomass, tree height or
stem density \citep{Piponiot2016, Rutishauser2016}. Some of the
determinants of post-disturbance biomass trajectories are already
identified, like the structure and composition of the pre-disturbance
community, or the post-disturbance environmental parameters
\citep{Herault2018}. Community diversity and composition trajectories,
however, have not been as thoroughly understood
\citep{Guitet2018, Molino2001}, and manifold biodiversity trajectories
might emerge given the variety of species response to disturbance and
the diversity of tropical forests
\citep{Lindenmayer2012, Garcia_florez2017}.

An early conceptual basis of the linkage between biodiversity and
disturbance is the Intermediate Disturbance Hypothesis (IDH). The IDH
states a relationship between the community diversity and the intensity
and frequency of disturbance events, and postulates a diversity peak at
intermediate level of disturbance \citep{Connell1978}. This is based on
the fluctuations of community environment following disturbance that
foster both competitively superior species and fast colonizers, and
prevents competitive exclusion \citep{Shea2004, Pulsford2016}. In
tropical forests, however, observations of the IDH often diverge from
theoretical expectations \citep{Hugues2007, Sheil2003, Norden2017}, and
the underlying processes might be complicated by the diversity of
tropical tree communities \citep{Lindenmayer2012, Garcia_florez2017}. In
this context, the IDH is controversial in tropical forests and remains
to be tested \citep{Hubbell2001, Fox2013, Sheil2013}.

Analysing community response to disturbance and grasp all aspects of
community changes requires a wide array of metrics
\citep{Sheil2003, Shea2004, Mayfield2010}. The analysis should first
consider community composition, which is crucial for conservation issues
and reveals the pool of species fostered or hampered by disturbance
\citep{Lavorel2002, Bellwood2006}. This should be completed with
diversity metrics encompassing both community richness and evenness to
assess the changes in community abundance distribution. Besides,
functional approaches have been shown to usefully complement pure
taxonomic approaches as they shed light on the species biological
attributes directly linking community diversity, composition, and
redundancy to ecosystem functioning \citep{Violle2007b, Baraloto2012a}.
In that respect, a vast literature allowed recognizing major traits
representing species ecological strategy and determining how they
respond to changing conditions \citep{Diaz2005}. Specifically, in
tropical forests, the functional approach revealed the emergence of
deterministic processes following disturbance. Such deterministic
processes entailed a shift from a dominance of ``conservative''
slow-growing species dealing with scarce resources, to a dominance of
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \citep{Rees2001, Reich2014, Herault2011}. This shift
is translated into the trajectories of average community value of key
functional traits related to resource acquisition, as leaf and stem
traits, and life-history strategy, as seed mass and maximum size
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}.

The functional approach also encompasses the analysis of functional
redundancy, that quantifies the amount of shared trait values among
species \citep{Carmona2016}. The typical high functional redundancy of
hyper-diverse tropical forests \citep{Bellwood2006} mitigates the
impacts of species removal on ecosystem functioning, and determines
community resilience after disturbance \citep{Elmqvist2003, Diaz2005}.

Here, we monitored over 30 years the response of 75 ha of Neotropical
forest plots set up on a gradient of disturbance intensity, from 10 to
60\% of above-ground biomass (AGB) loss. We made use of a large
functional traits database encompassing major leaf, stem, and
life-history traits in order to draw the taxonomic and functional
trajectories in terms of richness, evenness, composition, and
redundancy. Specifically, (i) we drew taxonomic and functional
post-disturbance trajectories and examined the underlying ecological
processes, (ii) we discussed the scope of the IDH regarding taxonomic
and functional facets of community diversity, and (iii) we analyzed
community resilience and time to recovery.

@@

Paracou station in French Guiana (5° 18'N and 52° 53'W) is located in a
lowland tropical rain forest in a tropical wet climate with mean annual
temperature of 26° C, and mean annual precipitation averaging 2980
mm.y\textsuperscript{-1} (30-y period). The climate comprises a 3-month
dry season (\textless{} 100 mm.month\textsuperscript{-1}) from
mid-August to mid-November, and a one-month dry season in March
\citep{Wagner2011}. Across all plots, elevation ranges from 5 to 50 m,
and the topography mainly corresponds to hilltops or hillsides, while
bottomlands cover less than 1 \% of the area. Plots are shallow
ferralitic acrisols over a layer of transformed saprolite with low
permeability and lateral drainage. Soil conditions are homogeneous, to
the exception of the highest hilltops where the thick surface allows a
free vertical drainage \citep{Gourlet-Fleury2004}.

The experiment is a network of twelve 6.25 ha plots (Table
\ref{tab:Tab1}) that underwent three disturbance treatments in 1987
according to a randomized plot design \citep{Gourlet-Fleury2004}.\\
The experiment comprised three replicates of three sylvicultural
treatments (hereafter plots T1, T2, and T3), and three control plots
(T0). All treatments T1, T2, and T3 comprised the logging of 10 trees/ha
with 50 cm minimum DBH that belonged to a set of 58 commercially
exploited species \citep{Gourlet-Fleury2004}. Treatment T2 additionally
comprised a thinning treatment by poison-girdling of non-commercially
exploited species randomly selected with an average of 30 trees/ha with
40 cm minimum DBH. Treatment T3 additionally comprised the logging of 15
trees/ha with 40 cm minimum DBH, and the poison-girdling of 20 trees/ha
with a 50 cm minimum DBH, all belonging to non-commercially exploited
species. Considering the silvicultural treatments and the following
damage, disturbance intensity was measured as the percentage of
aboveground biomass (\%AGB) lost between the first inventory in 1984 and
five years after disturbance \citep{Piponiot2016} estimated with the
BIOMASS R package \citep{Rejou2017}. The three treatments were then
transformed into a continuous disturbance intensity gradient with
increasing above-ground biomass (AGB) loss.

@@

\caption{\label{tab:Tab1}Intervention table, summary of the disturbance intensity for the 4 plot treatments in Paracou. Treatment intensities are defined by the minimum logging DBH (Diameter at Breast Height), the type of logged species (commercially exploited or not), the density of logged trees, and the total AGB (Above Ground Biomass) lost after treatment.}
\centering
\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
\toprule
Treatment & Timber & Thinning & Fuelwood & \%AGB lost\\
\midrule
Control & - & - & - & 0\\
T1, low & DBH $\geq$ 50 cm, commercially exploited species, $\approx$ 10   $trees.ha^{-1}$ & - & - & $[12-33]$\\
T2, intermediate & DBH $\geq$ 50 cm, commercially exploited species, $\approx$ 10  $trees.ha^{-1}$ & DBH $\geq$ 40 cm, non-commercially exploited species, $\approx$ 30   $trees.ha^{-1}$ & - & $[33-56]$\\
T3, high & DBH $\geq$ 50 cm, commercially exploited species, $\approx$ 10  $trees.ha^{-1}$ & DBH $\geq$ 50 cm, non-commercially exploited species, $\approx$ 15  $trees.ha^{-1}$ & 40 cm $\leq$ DBH $\leq$ 50 cm, non-commercially exploited species,\ $\approx$ 15 $trees.ha^{-1}$ & $[35-56]$\\
\bottomrule
\end{tabu}
\end{table}

@@

The study site corresponds to a tropical rainforest typical of the
Guiana Shield with a dominance of \emph{Fabaceae},
\emph{Chrysobalanaceae}, \emph{Lecythidaceae}, and \emph{Sapotaceae}. In
the 12 experimental plots, all trees above 10 cm DBH have been mapped
and measured annually since 1984. Trees are first identified with a
vernacular name assigned by the forest worker team, and afterward with a
scientific name assigned by botanists during regular botanical
campaigns. In 1984, specific vernacular names were given to 62 commmon
or commercially exploited species. More infrequent species were
identified under general identifiers only distinguishing trees and
palms. From 2003, botanical campaigns have been conducted every 5 to 6
years to identify all trees at the species level. In 2015, however,
identification levels still varied among plots and campaigns.

This variability of protocols in time raised methodological issues as
vernacular names usually correspond to various botanical species. This
resulted in significant taxonomic uncertainty that had to be accounted
for in the measure of composition and diversity metrics. Uncertainty
propagation was implented withing a Bayesian framework using
vernacular/botanical names associations to reconstitute complete
inventories at genus level from real incomplete ones. Vernacular names
were replaced through multinomial trials based on the association
probability \(\big[\alpha_1, \alpha_2,..., \alpha_V\big]\) observed
across all inventories between each vernacular name \emph{v} and all
species \(\big[s_1, s_2,..., s_N\big]\):

@@

Six functional traits representing the leaf economics (\emph{i.e.} leaf
thickness, toughness, total chlorophyll content, and specific leaf
area), and the stem economics (\emph{i.e.} wood specific gravity and
bark thickness) were obtained from the BRIDGE project \footnote{http://www.ecofog.gf/Bridge/}.
Trait values were assessed from a selection of individuals located in
nine permanent plots in French Guiana, including two in Paracou, and
comprised 294 species belonging to 157 genera. Whenever a species was in
the dataset but missed some trait values (10\% of the species), missing
values were filled using multivariate imputation by chained equation
\citep{Mice2011}. To account for the phylogenetic signal in the filling
process, imputations were based on samples of species from the same
genus or from the same family. Whenever a species was not in the
dataset, it was attributed a set of trait values randomly sampled among
species of the next higher taxonomic level (same genus or family). Two
life-history traits, maximum specific height and seed mass, came from
the Mariwenn database \footnote{https://www.ecofog.gf/mariwenn/}. The
database compiles information from a vast literature on the flora of
French Guiana \citep{Ollivier2007} and comprises 362 species belonging
to 188 genera. As seed mass information was classified into classes, no
data filling process was applied and analyses were restricted to the
botanical species recorded.

@@

Because of the variable precision of botanical identification efforts,
we had to conduct the taxonomic composition and diversity analysis at
the genus level. Taxonomic and functional trajectories of community
composition were drawn in a two-dimensional NMDS ordination plane. Two
NMDS using abundance-based (\emph{i.e.} Bray-Curtis) dissimilarity
measures were conducted to map either taxonomic or functional
composition, the latter based on the seven leaf, stem, and life history
traits (without seed mass classes). Trajectories along time were
reported through the Euclidean distance between the target inventories
and the 1984 pre-disturbance inventories of reference. Univariate
trajectories of the leaf, stem, and life-history traits were also
visualized with the community weighted means (CWM) \citep{Diaz2007}.
Species seed mass were given in 5 mass classes, and seed mass
trajectories were reported as the proportion of each class in the
inventories (Supplementary materials).

The taxonomic and functional trajectories were analysed from the 1984
pre-disturbance inventories of reference. The taxonomic diversity was
reported through species richness and the Hill number translation of the
Simpson index \citep{Hill1973}. The comparison between these two metrics
assesses community taxonomic richness and evenness: thereafter, results
will be discussed directly in terms of taxonomic richness and evenness.
Both indices are recommended for diversity studies \citep{Marcon2015},
and belong to the set of HCDT or generalized entropy corresponding,
respectively, to the 0 and 2 order of diversity (q). The functional
diversity was reported using the functional richness and functional
evenness, through the Rao index of quadratic entropy. The Rao index
combines species abundance distribution, and the average pairwise
functional dissimilarity between species computed by the Gower distance.

The impacts of the initial disturbance levels on the maximum gain or
loss in richness and evenness were tested with the Spearman rank
correlation tests. Richness and evenness trajectories were besides
analyzed through polynomial regression between (i) taxonomic and
functional richness and evenness, and (ii) the initial \%AGB loss at 10,
20, and 30 years after disturbance.

Finally, functional redundancy was measured as the overlap among species
in community functional space \citep{Carmona2016}. First, the
individuals of the trait database were mapped in the plane of the first
two axes from a PCA analysis. The PCA analysis lowered the weight of
correlations among traits as the axes are combinations of the most
decoupled traits. For each species, the traits probability density (TPD)
were computed from the mapping of individuals through two-dimension
kernel density estimators. Second, for each community, the TPD weigthed
by species abundance were summed accross the functional space. Third,
the functional space was divided into a 100 x 100 grid, and the number
of species with a positive TPD was counted in each cell. The average
count across cells minus 1 returned the Community Functional Redundancy,
which was the average number of species in the community that share the
same trait values.

@@

From 1984, the first pre-disturbance inventory, to 2015, 28 years after
disturbance, 828 388 individual trees and 591 botanical species spanning
223 genera and 64 families were recorded.

In undisturbed plots, both taxonomic and functional composition remained
stable (Fig. \ref{fig:NMDSplans}). In disturbed plots, both trajectories
followed marked and consistent trajectories over time.

The functional composition trajectory resembled, in disturbed
communities, to cyclic compositional changes with an incomplete recovery
of the initial composition (Fig.\ref{fig:NMDSplans}). The maximum
dissimilarity with the initial state was positively correlated with the
disturbance intensity for both taxonomic and functional composition
(\(\rho_{Spearman}^{Taxonomic}=0.87\) and
\(\rho_{Spearman}^{Functional}=0.90\), respectively). The maximum
dissimilarity with the initial state was reached for taxonomic
composition between 15 to 25 years, and between 20 to 25 years for
functional composition.

@@

\caption{Plot trajectories in terms of taxonomic composition (\textbf{(a)} and \textbf{(c)}), and functional composition (\textbf{(b)} and \textbf{(d)}) in a two-dimensional NMDS plane. Lower panels (\textbf{(c)} and \textbf{(d)}) represent the Euclidean distance to initial condition along the 30 sampled years. Shaded areas are the credibility intervals.}\label{fig:NMDSplans}
\end{figure*}

In control plots, Community Weighted Means (CWM) of functional traits
remained stable in time.\\
In disturbed plots, they mostly followed unimodal trajectories, either
stabilizing or returning towards their initial values, to the exception
of leaf chlorophyll content, which continued to increase 30 years after
disturbance for 4 out of 6 highly disturbed plots. Maximum height at
adult stage (\emph{Hmax}), leaf toughness, and wood specific gravity
(\emph{WSG}) decreased in time and then slightly increased, but remained
significantly lower than their initial value (Fig. \ref{fig:CWM}). Bark
thickness and specific leaf area (\emph{SLA}) both increased along time.
Bark thickness remained substantially high after 30 years, and
\emph{SLA} had almost recovered to its initial value. Whatever the
functional traits, the maximum difference to initial value was highly
correlated to the disturbance intensity. Positive correlations were
observed for leaf thickness, chlorophyll content, SLA and bark thickness
(\(\rho_{Spearman}^{Leaf thickness}=0.76\),
\(\rho_{Spearman}^{Chlorophyll content}=0.60\),
\(\rho_{Spearman}^{SLA}=0.93\),
\(\rho_{Spearman}^{Bark thickness}=0.71\)). Negative correlation was
observed for leaf toughness, WSG, and Hmax
(\(\rho_{Spearman}^{Leaf toughness}=-0.53\),
\(\rho_{Spearman}^{WSG}=-0.75\), \(\rho_{Spearman}^{Hmax}=-0.40\)) The
proportions of the three lightest seed mass classes increased in all
disturbed plots. After 30 years the proportion of lightest seed mass
class decreased while it stabilized for the two other lightest seed mass
classes (Supp. Mat. - Fig. S2).

@@

AM, EM \& BH designed the study, developed the analysis framework, and
interpreted the results. AM wrote the manuscript with contributions by
EM \& BH. All authors gave final approval for publication.
